Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-18 paper_io_sim.tex
minor edits

        The two main concerns of simulation are accuracy,
        the ability to faithfully reproduce real-world executions, and
        scalability, the ability to simulate large/long real-world
        executions quickly and with low RAM footprint. The above
        frameworks achieve different compromises between the two.  At
        one extreme are discrete-event models that capture
        ``microscopic'' behaviors of hardware/software systems (e.g.,
        packet-level network simulation, block-level disk simulation,
        cycle-accurate CPU simulation), which favor accuracy over
        speed.  At the other extreme are analytical models that capture
        ``macroscopic'' behaviors via mathematical models.  While these
        models lead to fast simulation, they must be developed
        carefully if high levels of accuracy are to be
        achieved~\cite{velhoTOMACS2013}.

@@

            This prototype does not simulate  bandwidth sharing and thus does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The Python 3.7 source code of this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}.

@@

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we ran \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we ran
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes had 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we used the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. As is commonly configured in HPC
        environments to avoid data loss, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback. NFS client and server read caches were enabled. 
        Therefore, all the writes happened at disk bandwidth, but
        reads can benefit from cache hits.

@@

        The page cache simulation model drastically reduced I/O simulation
        errors in each application task (Figure~\ref{fig:single_error}). The first read was not impacted
        as it only involved uncached data. Errors were reduced from an average
        of 345\% in the original \wrench to 46\% in the Python prototype and
        39\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimated read and write times, due to the lack
        of page cache simulation. Results with files of 50~GB and 75~GB
        showed similar behaviors and were not reported for brevity.

@@

        Simulated memory profiles were highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory profiles almost exactly matched the
        real ones, although dirty data seemed to be flushing faster in real
        life than in simulation \tristan{any idea why?}, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reached
        total memory during the first write, triggering dirty data
        flushing, and droped back to cached memory when application tasks
        released anonymous memory. Simulated cached memory was highly
        consistent with real values, except toward the end of Read 3 where
        it slightly increased in simulation but not in reality. This
        occurred due to the fact that after Write 2, File 3 was only partially
        cached in simulation whereas it was entirely cached in the real
        system. In all cases, dirty data remained under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibited nearly
        identical memory profiles, which reinforces the confidence in our
        implementations.

        The content of the simulated memory cache was also highly
        consistent with reality (Figure~\ref{fig:single_cache}). With 20~GB
        files, the simulated cache content exactly matched reality, since
        all files fit in page cache. With 100~GB files, a slight
        discrepancy was observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real
        execution indeed, File 3 was entirely cached after Write 2, whereas
        in the simulated execution, only a part of it was cached. This was
        due to the fact that the Linux kernel tends to not evict pages that
        belong to files being currently written (File 3 in this case),
        which we could not easily reproduce in our model.

            \begin{figure*}
                \begin{subfigure}{\linewidth}
                    \centering
                    \includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
                \end{subfigure}
                \caption{Concurrent results with 3~GB files (\textit{Exp 2}). Averages on 5 repetitions.}
                \label{fig:multi_local}
            \end{figure*}
        \subsection{Concurrent applications (Exp 2)}

@@

            \begin{figure*}
                \begin{subfigure}{\linewidth}
                    \centering
                    \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
                \end{subfigure}
                \caption{NFS results with 3~GB files (\textit{Exp 3}). Averages on 5 repetitions.}
                \label{fig:multi_nfs}
                \end{figure*}

>>>>

        The two main concerns for simulation are accuracy,
        the ability to faithfully reproduce real-world executions, and
        scalability, the ability to simulate large/long real-world
        executions quickly and with low RAM footprint. The above
        frameworks achieve different compromises between the two.  At
        one extreme are discrete-event models that capture
        ``microscopic'' behaviors of hardware/software systems (e.g.,
        packet-level network simulation, block-level disk simulation,
        cycle-accurate CPU simulation), which favor accuracy over
        speed.  At the other extreme are analytical models that capture
        ``macroscopic'' behaviors via mathematical models.  While these
        models lead to fast simulation, they must be developed
        carefully if high levels of accuracy are to be
        achieved~\cite{velhoTOMACS2013}.

@@

            This prototype does not simulate  bandwidth sharing and thus does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The Python 3.7 source code is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}.

@@

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp~1}), we ran \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp~2}), we ran
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes had 32 CPU
        cores.
        In the third experiment (\textit{Exp~3}), we used the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. As is commonly configured in HPC
        environments to avoid data loss, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback. NFS client and server read caches were enabled. 
        Therefore, all the writes happened at disk bandwidth, but
        reads could benefit from cache hits.

@@

        The page cache simulation model drastically reduced I/O simulation
        errors in each application task (Figure~\ref{fig:single_error}). The first read was not impacted
        as it only involved uncached data. Errors were reduced from an average
        of 345\% in the original \wrench to 46\% in the Python prototype and
        39\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimated read and write times, due to the lack
        of page cache simulation. Results with files of 50~GB and 75~GB
        showed similar behaviors and are not reported for brevity.

@@

        Simulated memory profiles were highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory profiles almost exactly matched the
        real ones, although dirty data seemed to be flushing faster in real
        life than in simulation, \tristan{any idea why?} a behavior also
        observed with 100~GB files. With 100~GB files, used memory reached
        total memory during the first write, triggering dirty data
        flushing, and droped back to cached memory when application tasks
        released anonymous memory. Simulated cached memory was highly
        consistent with real values, except toward the end of Read 3 where
        it slightly increased in simulation but not in reality. This
        occurred due to the fact that after Write 2, File 3 was only partially
        cached in simulation whereas it was entirely cached in the real
        system. In all cases, dirty data remained under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibited nearly
        identical memory profiles, which reinforces the confidence in our
        implementations.

        The content of the simulated memory cache was also highly
        consistent with reality (Figure~\ref{fig:single_cache}). With 20~GB
        files, the simulated cache content exactly matched reality, since
        all files fitted in page cache. With 100~GB files, a slight
        discrepancy was observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real
        execution indeed, File 3 was entirely cached after Write 2, whereas
        in the simulated execution, only a part of it was cached. This was
        due to the fact that the Linux kernel tends to not evict pages that
        belong to files being currently written (File 3 in this case),
        which we could not easily reproduce in our model.

            \begin{figure*}
                \begin{subfigure}{\linewidth}
                    \centering
                    \includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
                \end{subfigure}
                \caption{Concurrent results with 3~GB files (\textit{Exp 2})}
                \label{fig:multi_local}
            \end{figure*}
        \subsection{Concurrent applications (Exp 2)}

@@

            \begin{figure*}
                \begin{subfigure}{\linewidth}
                    \centering
                    \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
                \end{subfigure}
                \caption{NFS results with 3~GB files (\textit{Exp 3})}
                \label{fig:multi_nfs}
                \end{figure*}
